Calculating Biological Module Enrichment or Depletion and Visualizing Data on Large-scale Molecular Maps with \pkg{ACSNMineR} and \pkg{RNaviCell} packages

### 2016-08-31 jss/Deveau.tex
[skip ci] text update

The dataset corresponding to this study is available as a Bioconductor package.
The code shown below is creating a list of differentially expressed genes
between ER positive and ER negative samples, and calculates the enrichment in
ACSN maps from this list of genes. As seen in Table \ref{tab:table_mainz},
there is one map (Cell Cycle) and six modules (belonging to the Cell Cycle, DNA
repair and EMT motility maps) enriched.

@@

\begin{table}[h!]
  \centering
  \caption{ACSN maps enrichment for genes differentially expressed between ER
positive and ER negative samples in breast cancer.  Module : name of the
module. Mod. size: size of the module. Nb genes: number of genes from input
which are found in the module. pval: raw p-value. Cor. pval: corrected p-value.}
\label{tab:table_mainz}

@@

The Figure \ref{fig:mainz} is displaying the map for gene that having a
corrected p-value $< 0.05$ and a log fold-change $>1$. we can see that the genes
are concentrated in the regions of the map corresponding to the cell cycle, cell
motility, apoptosis and survival.

@@

We also describe in this work the \pkg{RNaviCell} package, a R package
convenient to use with \pkg{ACSNMineR}. This package is dedicated to create
web-based and interactive data visualization on ACSN maps. Users can use this
tools to represent genes of interest that have been shown to be related to the
maps by calculating enrichment with \pkg{ACSNMineR}.  Creating maps with the
graphical user interface of the ACSN website can be a tedious task if the user
has multiple samples or gene lists, and wants to compare their representations
on ACSN maps. The \pkg{RNaviCell} package can be used to automate the process
of creating the graphical representations automatically.  The maps are
displayed in a browser and are interactive, with the possibility for the user
to zoom in and out, search for genes or molecular species, and see the details
of the molecular reactions (what partners are involved, what is the state of a
given species, etc.). For more details on how to use the interface and the
different possibilities, see \cite{kuperstein2013navicell},
\cite{bonnet2015navicell} and \cite{kuperstein2015atlas}.

We have shown how the packages \pkg{ACSNMineR} and \pkg{RNaviCell} can be
combined to analyze expression data from breast cancer samples, and also to
analyze the frequency of mutated genes in glioblastoma cancer samples.

>>>>

The dataset corresponding to this study is available as a Bioconductor package.
The code shown below is creating a list of differentially expressed genes
between ER positive and ER negative samples, and calculates the enrichment in
ACSN maps from this list of genes. As seen in Table \ref{tab:table_mainz},
there is one map (DNA repair) and seven modules (belonging to the Cell Cycle, DNA
repair and Apoptosis maps) enriched.

@@

\begin{table}[h!]
  \centering
  \caption{ACSN maps enrichment for genes differentially expressed between ER
positive and ER negative samples in breast cancer.  Module : name of the
map/module. Mod. size: size of the module. Nb genes: number of genes from input
which are found in the module. pval: raw p-value. Cor. pval: corrected p-value.}
\label{tab:table_mainz}

@@

The Figure \ref{fig:mainz} is displaying the map for genes having a
corrected p-value $< 0.05$.

@@

We also describe in this work the \pkg{RNaviCell} package, a R package
convenient to use with \pkg{ACSNMineR}. This package is dedicated to create
web-based and interactive data visualization on ACSN maps. Users can use this
tools to represent genes of interest that have been shown to be related to the
maps by calculating enrichment with \pkg{ACSNMineR}.  Creating maps with the
graphical user interface of the ACSN website can be a tedious task if the user
has multiple samples or gene lists, and wants to compare their representations
on ACSN maps. The \pkg{RNaviCell} package can be used to automate the process
of creating the graphical representations automatically.  The maps are
displayed in a browser and are interactive, with the possibility for the user
to zoom in and out, search for genes or molecular species, and see the details
of the molecular reactions (what partners are involved, what is the state of a
given species, etc.). For more details on how to use the interface and the
different possibilities, see \cite{kuperstein2013navicell},
\cite{bonnet2015navicell} and \cite{kuperstein2015atlas}. We have shown how the
packages \pkg{ACSNMineR} and \pkg{RNaviCell} can be combined to analyze
expression data from breast cancer samples, and also to analyze the frequency
of mutated genes in glioblastoma cancer samples.
